Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.Rmd
responses 1.2, 1.3, 1.4, 2.2, 2.8
not happy with the blurred jittering lines

---
title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #or none
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: blue
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---

```{r eval=TRUE, echo=FALSE}
# setwd("paper/PaperCEUS/") # knit directory: Document
# system("latexdiff PaperCEUS.tex PaperCEUS_rev1.tex > tracked_changes_rev1.tex")
# system("pdflatex tracked_changes_rev1.tex")
```

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      # root.dir = "paper/PaperCEUS/",
                      out.extra = "")  # This forces knitr to label all figures.

@@

To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].

@@

# od_all = readRDS("paper/paperTRA/load/od_all.Rds")
# od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
# 
# par(mfrow = c(1, 2))
# 
# odcombined = rbind(
#   od_all |>
#     transmute(Total, Disaggregation = " Original"),
#   od_jittered_filter |>
#     transmute(Total, Disaggregation = "Jittered")
# )
# odcombined |>
#   sample_n(10000, weight = odcombined$Total) |>
#   mutate(Total = Total / 1000) |>
#   ggplot() +
#   geom_sf(aes(alpha = Total)) +
#   facet_wrap(~Disaggregation) +
#   theme_void() +
#   theme(legend.position="none",
#         strip.text.x = element_blank()) +
#   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
# 
# ggsave(filename = "img/jitter_rev2.png", plot = last_plot(), width = 1000, height = 400, units = "px", dpi = 300, type = "cairo")

knitr::include_graphics("img/jitter_new.png", error = FALSE)
```

@@

<!-- Include a map with the legs of journeys, and which modes are replacing which, and what impacts are bein estimated at each one. See PTcar_legs.R -->

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles. 
For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

knitr::kable(summary1,
      digits = 0,
      format.args = list(big.mark = " "),
      col.names = c("Target", "Routing", "Total trips",
                    "Baseline Cycling + PT", "Potencial Cycling + PT"),
      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(4:5), width = "7em")

@@

knitr::kable(summary21 |> select(-Baseline),
      digits = 0,
      format.args = list(big.mark = " "),
      # col.names = c("Target", "Routing", "Total",
      #               "Bus", "Ferry", "Train", "Tram"),
      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(1:7), width = "4.5em")

@@

Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.

>>>>

---
title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #ou nenhum
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: black
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---
\hypersetup{citecolor=black}

```{r eval=FALSE, echo=FALSE}
# setwd("paper/PaperCEUS/") # knit directory: Document
system("latexdiff paper/PaperCEUS/PaperCEUS_prerev.tex paper/PaperCEUS/PaperCEUS_rev1.tex > paper/PaperCEUS/PaperCEUS_diff1.tex")
#and change in tex the blue for another color (green? 0,0.5,0) and remove \uwave
# \definecolor{green}{rgb}{0,0.5,0}
# {\protect\color{green}{#1}}}
#and compile PDF!
system("pdflatex paper/PaperCEUS/PaperCEUS_diff1.tex")
```

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      # root.dir = "paper/PaperCEUS/", # knit directory: Document, instead of the default
                      out.extra = "")  # This forces knitr to label all figures.

@@

The Portuguese Government has established a national cycling strategy, which outlines targets for the percentage of trips to be made by bicycle in urban areas.
Specifically, these targets aim for 4% of all trips in 2025 and 10% by 2030 [@ENMAC].
Moreover, the strategy emphasizes that this increase should be achieved by substituting car trips for bicycle trips, as a way to reduce the environmental and social impacts of car use.

To support the implementation of this strategy, the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].

@@

# od_all = readRDS("paper/paperTRA/load/od_all.Rds")
# od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
# 
# odcombined = rbind(
#   od_all |>
#     transmute(Total, Disaggregation = " Original"),
#   od_jittered_filter |>
#     transmute(Total, Disaggregation = "Jittered")
# )
# 
# cairo_pdf(filename = "img/jitter_cairo.pdf", onefile= T, width=10, height=4) 
# 
# odcombined |>
#   sample_n(10000, weight = odcombined$Total) |>
#   mutate(Total = Total / 1000) |>
#   ggplot() +
#   geom_sf(aes(alpha = Total)) + # blured
#   facet_wrap(~Disaggregation) +
#   theme_void() +
#   theme(legend.position="none",
#         strip.text.x = element_blank()) +
#   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
# 
# dev.off()
# 
# png::writePNG(pdftools::pdf_render_page("img/jitter_cairo.pdf", dpi = 300), "img/jitter_cairo.png")

# ggsave(filename = "img/jitter_rev2.png", plot = last_plot(), width = 1000, height = 400, units = "px", dpi = 300, type = "cairo")

knitr::include_graphics("img/jitter_cairo.png", error = FALSE)
```

@@

<!-- Include a map with the legs of journeys, and which modes are replacing which, and what impacts are bein estimated at each one. See PTcar_legs.R -->
Following the clearly established national strategy, we estimated the socio-environmental potential impacts of shifting car trips by bicycle in combination with PT, in the LMA, for the two targets (4% and 10%).

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles. 
For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

knitr::kable(summary1,
      digits = 0,
      format.args = list(big.mark = " "),
      col.names = c("Target", "Routing", "Total trips",
                    "Baseline Cycling + PT", "Potencial Cycling + PT"),
      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(4:5), width = "7em")

@@

knitr::kable(summary21 |> select(-Baseline),
      digits = 0,
      format.args = list(big.mark = " "),
      # col.names = c("Target", "Routing", "Total",
      #               "Bus", "Ferry", "Train", "Tram"),
      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(1:7), width = "4.5em")

@@

Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions, our findings indicate a potential surge in demand (+6.3%) of public transport interfaces to align with the national strategies.
This added demand is conceptually driven from individuals utilising bicycles, either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system.
Additionally, they may need to explore options to enhance bicycle carriage capacity on public transport services, ensuring alignment with national targets.

This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT. This process involves a minimum of data requirements, such as the origin-destination data, road network, and the PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.
Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.tex
responses 1.2, 1.3, 1.4, 2.2, 2.8
not happy with the blurred jittering lines

@@

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
the Lisbon's Metropolitan Department of Transport commissioned
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network \citep{felix2023}.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter_new}

@@

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in Table
\ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{7em}>{\raggedleft\arraybackslash}p{7em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385\\
4\% & direct & 500 880 & 2 274 & 18 944\\
10\% & safe & 538 514 & 2 312 & 52 323\\
10\% & direct & 500 880 & 2 274 & 48 609\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.

>>>>

\hypersetup{citecolor=black}

@@

The Portuguese Government has established a national cycling strategy,
which outlines targets for the percentage of trips to be made by bicycle
in urban areas. Specifically, these targets aim for 4\% of all trips in
2025 and 10\% by 2030 \citep{ENMAC}. Moreover, the strategy emphasizes
that this increase should be achieved by substituting car trips for
bicycle trips, as a way to reduce the environmental and social impacts
of car use.

To support the implementation of this strategy, the Lisbon's
Metropolitan Department of Transport commissioned
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network \citep{felix2023}.

@@

{\centering \includegraphics[width=1\linewidth,]{img/jitter_cairo}

@@

Following the clearly established national strategy, we estimated the
socio-environmental potential impacts of shifting car trips by bicycle
in combination with PT, in the LMA, for the two targets (4\% and 10\%).

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in Table
\ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{7em}>{\raggedleft\arraybackslash}p{7em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385\\
4\% & direct & 500 880 & 2 274 & 18 944\\
10\% & safe & 538 514 & 2 312 & 52 323\\
10\% & direct & 500 880 & 2 274 & 48 609\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions,
our findings indicate a potential surge in demand (+6.3\%) of public
transport interfaces to align with the national strategies. This added
demand is conceptually driven from individuals utilising bicycles,
either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the
challenge of accommodating this increased demand for bicycle parking
and/or docks from a bike sharing system. Additionally, they may need to
explore options to enhance bicycle carriage capacity on public transport
services, ensuring alignment with national targets.

This analysis can be applied to other metropolitan areas, providing a
basis for decision-making and policy development to support a modal
shift from car to PT. This process involves a minimum of data
requirements, such as the origin-destination data, road network, and the
PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level
targets specific to those areas or consider the possibility of utilising
the same targets as outlined in the Portuguese national strategy, in
order to compare results with different areas. Nevertheless, the extent
of the potential shift from car to PT is highly dependent on the
distances typically travelled in such contexts, and the availability of
PT services - which can vary significantly between regions, and for
which we are not assuming any change in this study. This paper focus
primarily on the potential of making cycling and PT a viable alternative
to car trips, and the socio-environmental benefits of such a shift.
